Reasonably Programmable Literal Notation

### 2018-06-18 icfp18/syntax-icfp18.tex
fixed spelling of analogous

% In short, the problem is that library providers cannot reason modularly about syntactic determinism (because other extensions might define overlapping forms), and if the desugaring of the program text is held abstract, programmers can no longer reason about types, binding and segmentation (i.e. answer questions like those above.) This is burdensome at all scales, but particularly when programming in the large, where it is common to encounter a wide variety of libraries \cite{DBLP:conf/sac/LammelPS11}. 
Forcing the programmer to reason transparently to answer basic questions like these defeats the ultimate purpose of syntactic sugar: decreasing cognitive cost \cite{Green89}. Analagous problems do not arise when programming without syntax extensions in languages like ML --- programmers can reason lexically about where variables and other symbols are bound, and types mediate abstraction over function and module implementations \cite{B304}. Ideally, the programmer would be able to abstract in some analagous manner over the implementation of an unfamiliar notation. % From this perspective, it is unsurprising that the use of Camlp4 has been deprecated in the OCaml system.

@@

\subsubsection{Expansion Typing}
Having found the definition of the \li{$regex} TLM, the client can immediately determine the type of the expansion being generated at each application site because it is specified explicitly by the clause \li{at Regex.t} on Line~2 of Fig.~\ref{fig:regex-tlm-def}. The expansion type of a TLM is analagous to the return type of a function. The identity of \li{Regex.t} is determined relative to the TLM definition site, not at each application site, so the module \li{Regex} need not be in scope at the application site, or it can have been shadowed by a different module. % In Reason, TLM definitions can refer to exactly those modules that a Reason module at the top level of the defining package would have access to, i.e. other top-level modules in the package as well as modules exported by other packages that have been imported by the build system.

@@

Segment types are somewhat analagous to the argument types of a function. The difference is that the argument signature of a function is the same every time the function is applied, i.e. it is associated with the function itself, whereas the inferred segmentation can differ for each choice of literal body. This, of course, gives TLMs substantially more notational flexibility, even relative to infix or mixfix function notation (where the number of subexpressions is fixed \cite{wieland2009parsing}), as intended.% (even infix or mixfix functions \cite{wieland2009parsing}).

@@

% \subsection{Splicing}\label{sec:splicing-and-hygiene}
% When the parse function determines that some segment of the literal body is a spliced expression, according to whatever syntactic criteria it deems suitable, it can indirectly refer to it in the encoding it produces using the \li{SplicedE} constructor of \li{proto_expr}, which takes a value of type \li{segment} that indicates the zero-indexed location of the spliced expression relative to the start of the provided literal body. The \li{SplicedE} constructor also requires a value of type \li{proto_typ}, which indicates the type that the spliced expression is expected to have. Types can be spliced out by using the \li{SplicedT} constructor of \li{proto_typ} analagously.

@@

\begin{figure}[t]
\begin{subfigure}[t]{0.55\textwidth}
\begin{lstlisting}[morekeywords={expression}]
  module Lambda = {/* a typical encoding */};
  module LambdaNotation {
    notation $term at Lambda.term { 
      lexer SimpleLangLexer;
      expression parser LambdaParser.term_e;
      pattern parser LambdaParser.term_p;
      dependencies 
        { module Lambda = Lambda } };
    notation $v at Lambda.v {/* analagous */}
  };
\end{lstlisting}
\vspace{-6px}
\caption{Expression and pattern TLMs for \lismall{Lambda} terms and values}
\label{fig:lam}
\vspace{2px}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.42\textwidth}
\begin{lstlisting}
  open LambdaNotation;
  exception Unbound(Lambda.var);
  let rec eval = $term.(fun 
  | `(x)` => raise Unbound(x)
  | `(SURLlamEURL xSURL.EURLe)` => $v `(SURLlamEURL xSURL.EURLe)`
  | `(e1SURL(EURLe2SURL)EURL)` => {
    let $v `(SURLlamEURL xSURL.EURLe)` = eval(e1);
    let v2 = eval(e2);
    eval(`(SURL[EURLv2SURL/EURLxSURL]EURLe)`) 
  });
\end{lstlisting}
\vspace{-6px}
\caption{A reasonably elegant \lismall{Lambda} evaluator}
\label{fig:lam-eval}
\end{subfigure}
\vspace{2px}
\caption{Expression and pattern literal notation for lambda terms and values}
\vspace{-8px}
\label{fig:lam-example}
\end{figure}

@@

From a client programmer's perspective, reasoning principles analagous to those described in Sec. \ref{sec:client-perspective} for expression TLMs are available. \textbf{Responsibility} is assigned by the same protocol, and the type annotation on the responsible TLM governs the type of the generated pattern, satisfying the \textbf{Expansion Typing} condition. Patterns can contain module paths and type annotations, so pattern TLMs are governed by the same \textbf{Context Independence} condition as expression TLMs. The same protocol around \textbf{Segmentation} is also enforced.

@@

% \begin{figure}
% \begin{subfigure}[t]{0.51\textwidth}
% \begin{lstlisting}
%   module Lambda = {
%     type var = int;
%     type tm = Var(var) | Lam(var, tm) | Ap(tm, tm);
%     type v  = VLam(var, tm);
%     let subst : v -> var -> tm -> tm = (* ... *)
%   };
% \end{lstlisting}
% \caption{Lambda calculus terms}
% \end{subfigure}
% \hfill
% \begin{subfigure}[t]{0.49\textwidth}
% \begin{lstlisting}
%   module Lam = { /* a typical encoding */ };
%   module LamNotation {
%     notation $term at Lam.term { 
%       lexer SimpleLangLexer;
%       expression parser LamParser.tm_expr;
%       pattern parser Lam.tm_pat;
%       dependencies { module Lam }
%     }
%     notation $v at Lam.v { /* analagous */ } 
%   };
% \end{lstlisting}
% \vspace{-4px}
% \caption{Lambda TLM definitions}
% \vspace{2px}
% \end{subfigure}
% \caption{Lambda calculus example}
% \end{figure}

@@

% To maintain the abstract binding discipline, variable patterns can appear only within spliced patterns. Enforcing this restriction is straightforward: we simply have not defined a variant of the \li{proto_pat} type that encodes variable patterns (wildcards are allowed.)  This restriction ensures that only variables visible to the client in a spliced pattern are bound in the corresponding branch expression. This is analagous to the capture avoidance principle for expression TLMs.

@@

Instead, we define \emph{unexpanded typing contexts}, $\uGamma$, as pairs of the form $\uGG{\uG}{\Gamma}$, where $\uG$ maps each expression identifier $\ux \in \domof{\uG}$ to a variable, $x$, written $\vExpands{\ux}{x}$. The typing context, $\Gamma$, only tracks the type of these variables. We define the identifier update operation $\ctxUpdate{\uG}{\ux}{x}$ as producing the expression identifier expansion context that maps $\ux$ to $x$, written $\vExpands{\ux}{x}$, and defers to $\uG$ for all other expression identifiers, with no requirement that $\ux$ be apart from $\domof{\uG}$. %Note the distinction between update and extension (which requires that the new identifier is not already in the domain.) %We write $\uGammaOK{\uGamma}$ when $\uGamma=\uGG{\uG}{\Gamma}$ and each expression variable in $\uG$ is assigned a type by $\Gamma$.
%\begin{definition} $\uGammaOK{\uGG{\uG}{\Gamma}}$ iff for each $\vExpands{\ux}{x} \in \uG$, we have $\Ghyp{x}{\tau} \in \Gamma$ for some $\tau$.\end{definition}
%\noindent 
We define $\uGamma, \uGhyp{\ux}{x}{\tau}$ when $\uGamma = \uGG{\uG}{\Gamma}$ as an abbreviation of $\uGG{\ctxUpdate{\uG}{\ux}{x}}{\Gamma, \Ghyp{x}{\tau}}$. Unexpanded type formation contexts, $\uDelta$, are analagous.

@@

% \emph{Unexpanded type formation contexts}, $\uDelta$, consist of a \emph{type identifier expansion context}, $\uD$, paired with a standard type formation context, $\Delta$, and operate analagously (see supplement.)
% of the form $\uDD{\uD}{\Delta}$, i.e. they . We similarly define $\uDelta, \uDhyp{\ut}{t}$ when $\uDelta=\uDD{\uD}{\Delta}$ as an abbreviation of $\uDD{\ctxUpdate{\uD}{\ut}{t}}{\Delta, \Dhyp{t}}$.%type identifier expansion context is always extended/updated together with

@@

Rule \textsc{ee-ap-sptlm} for pattern TLM definitions shown in Fig. \ref{fig:ee-def-setlm} is analagous but simpler, because in $\miniVersePat$ patterns are entirely structural (there are no module paths that they might depend on). %As such, we do not need the machinery for dependency passing.
% {\vspace{-3px}\small\begin{mathpar}
% \inferrule[ee-def-sptsm]{
%   \expandsTU{\uDelta}{\utau}{\tau}\\
%   \hastypeU{\emptyset}{\emptyset}{\eparse}{\aparr{\tBody}{\tParseResultPat}}\\\\
%   \evalU{\eparse}{\eparse'}\\
%   \expandsUP{\uDelta}{\uGamma}{\uPsi}{\uPhi, \uPhyp{\tsmv}{a}{\tau}{\eparse'}}{\ue}{e}{\tau'}
% }{
%   \expandsUPX{\usyntaxup{\tsmv}{\utau}{\eparse}{\ue}}{e}{\tau'}
% }
% \end{mathpar}}
% \vspace{-3px}

@@

The typed pattern expansion rule governing pattern TLM application, Rule~\textsc{pe-ap-sptlm} in Fig.~\ref{fig:ee-ap-setlm}, is analagous but again simpler because there is no need to pass dependencies into patterns in $\miniVersePat$.\footnote{If there were a need to pass dependencies into patterns, we would need either pattern functions or to use an encoding trick, e.g. by requiring the result be a function containing a ``dummy'' match with one pattern.} We describe proto-pattern validation in Sec.~\ref{sec:s-PEV} below.
% {\small\begin{mathpar}
% \inferrule[pe-ap-sptsm]{
%   \encodeBody{b}{\ebody}\\
%   \evalU{\ap{\eparse}{\ebody}}{\aein{\mathtt{SuccessP}}{\ecand}}\\
%   \decodeCEPat{\ecand}{\cpv}\\\\
%   \segOK{\segof{\cpv}}{b}\\
%   \cvalidP{\upctx}{\pscene{\uDelta}{\uPhi}{b}}{\cpv}{p}{\tau}
% }{
%   \patExpands{\upctx}{\uPhi', \uPhyp{\tsmv}{a}{\tau}{\eparse}}{\utsmap{\tsmv}{b}}{p}{\tau}
% }
% \end{mathpar}}

@@

The rule for references to spliced unexpanded types and patterns are analagous (see supplement). Note that we did not give examples of type splicing in the previous sections, but it is occasionally useful. For example, a TLM that implements a parser generator in the style of Menhir could need type splicing for reading in the non-terminal types, e.g. as on Line 9 of Fig.~\ref{fig:regex-parser}.
% \noindent
% We first splice out the requested segment. The second premise expands the type annotation under an empty context, because the type annotation must be meaningful at the application site (so, independent of $\Delta$ and $\Gamma$) and not itself make any assumptions about the application site context. Spliced types can appear in the annotation. The third premise performs typed expansion of the spliced unexpanded expression under the application site contexts, but not the expansion-local contexts. The final two premises ensure that these contexts are disjoint, again to force capture avoidance.

% The rule for references to spliced unexpanded patterns is entirely analagous (see supplement).
% {\small\begin{mathpar}
% \inferrule[ppv-spliced]{
%   \parseUPat{\bsubseq{b}{m}{n}}{\upv}\\
%   \cvalidT{\emptyset}{\tsceneUP{\uDelta}{b}}{\ctau}{\tau}\\
%   \patExpands{\upctx}{\uPhi}{\upv}{p}{\tau}
% }{
%   \cvalidP{\upctx}{\pscene{\uDelta}{\uPhi}{b}}{\acesplicedp{m}{n}{\ctau}}{p}{\tau}
% }
% \end{mathpar}}

@@

% \endgroup
The additional second clause simply states that the final expansion produced by proto-expression validation is well-typed under the combined application site and expansion-internal context (because spliced terms are distinguished only in the proto-expansion, but not in the final expansion). % The combined context can only be formed if these are disjoint.
The proof proceeds by mutual rule induction and appeal to the analagous typed pattern expansion theorem and simple lemmas about type expansion   and proto-type validation. The only issue is that it is not immediately clear that the mutual induction is well-founded, because the case in the proof of part 2 for Rule \textsc{pev-spliced} invokes part 1 of the induction hypothesis on a term that is not a sub-term of the conclusion, but rather parsed out of the literal body, $b$. To establish that the mutual induction is well-founded, then, we need to explicitly establish a decreasing metric. The intuition is that parsing a term of out a literal body cannot produce a bigger term than the term that contained that very literal body. 
% More specifically, {the sum of the lengths of the literal bodies that appear in the term strictly decreases each time you perform a nested TLM application} because some portion of the term has to be consumed by the TLM name and the delimiters. 
The details are given in the supplement.

@@

% % The following theorem summarizes the abstract reasoning principles available to programmers when applying a pattern TLM. Most of the labeled clauses are analagous to those described above, so we omit their descriptions.
% % % \begingroup
% % % \def\thetheorem{\ref{thm:spTLM-Typing-Segmentation}}
% % \begin{theorem}[spTLM Reasoning Principles]
% % % \label{thm:spTLM-Typing-Segmentation}
% % If $\patExpands{\upctx}{\uPhi}{\utsmap{\tsmv}{b}}{p}{\tau}$ where $\uDelta=\uDD{\uD}{\Delta}$ and $\uGamma=\uGG{\uG}{\Gamma}$ then all of the following hold:
% % \begin{enumerate}[noitemsep,nolistsep]
% %         \item (\textbf{Typing 1}) $\uPhi=\uPhi', \uPhyp{\tsmv}{a}{\tau}{\eparse}$ and $\patType{\pctx}{p}{\tau}$
% %         \item $\encodeBody{b}{\ebody}$
% %         \item $\evalU{\eparse(\ebody)}{\aein{\mathtt{SuccessP}}{\ecand}}$
% %         \item $\decodeCEPat{\ecand}{\cpv}$
% %         \item (\textbf{Segmentation}) $\segOK{\segof{\cpv}}{b}$
% %         \item $\segof{\cpv} = \sseq{\acesplicedt{n'_i}{m'_i}}{\nty} \cup \sseq{\acesplicedp{m_i}{n_i}{\ctau_i}}{\npat}$
% %         \item (\textbf{Typing 2}) $\sseq{
% %               \expandsTU{\uDelta}
% %               {
% %                 \parseUTypF{\bsubseq{b}{m'_i}{n'_i}}
% %               }{\tau'_i}
% %             }{\nty}$ and $\sseq{\istypeU{\Delta}{\tau'_i}}{\nty}$
% %         \item (\textbf{Typing 3}) $\sseq{
% %           \cvalidT{\emptyset}{
% %             \tsceneUP
% %               {\uDelta}{b}
% %           }{
% %             \ctau_i
% %           }{\tau_i}
% %         }{\npat}$ and $\sseq{\istypeU{\Delta}{\tau_i}}{\npat}$
% %         \item (\textbf{Typing 4}) $\sseq{
% %           \patExpands
% %             {\uGG{\uG_i}{\pctx_i}}
% %             {\uPhi}
% %             {\parseUPatF{\bsubseq{b}{m_i}{n_i}}}
% %             {p_i}
% %             {\tau_i}
% %         }{\npat}$  and $\sseq{\patType{\pctx_i}{p_i}{\tau_i}}{\npat}$
% %       \item (\textbf{Visibility}) $\uG = \biguplus_{0 \leq i < \npat} \uG_i$ and $\Gamma = \bigcup_{0 \leq i < \npat} \pctx_i$
% %         \begin{quote}
% %           \begin{grayparbox}
% %           The hypotheses generated by the TLM application are exactly those generated by the spliced patterns.
% %           \end{grayparbox}
% %         \end{quote}
% % \end{enumerate}
% % \end{theorem}
% % 
% % \begin{proof} 
% % The proof, in the supplement, relies on an auxiliary lemma about decomposing proto-patterns.

@@

% % % \item[{\text{(\ref{rule:cvalidE-U-tap})}}~\textbf{through}~{\text{(\ref{rule:cvalidE-U-case})}}] These cases follow analagously, i.e. we apply the IH, part 2 to all proto-expression validation judgements, Lemma \ref{lemma:candidate-expansion-type-validation} to all proto-type validation judgements, the identification convention to ensure that extended contexts remain disjoint, weakening and exchange as needed, and the corresponding typing rule in Rules (\ref{rule:hastypeU-tap}) through (\ref{rule:hastypeU-case}).
% % % \\

@@

% % % \item[\VExpof{\text{\ref{rule:hastypeU-tlam}}}~\text{through}~\VExpof{\text{\ref{rule:hastypeU-case}}}] These cases follow analagously, i.e. we apply the IH to all proto-expression validation premises, Lemma \ref{lemma:candidate-expansion-type-validation} to all proto-types validation premises, weakening and exchange as needed, and then apply the corresponding typing rule.
% % % \\

@@

% In contrast, when a programmer encounters, for example, a function call like the call to \li{read_data} on Line 3, the analagous questions can be answered by following clear protocols that become ``cognitive reflexes'' after sufficient experience with the language, even if the programmer has no experience with the library defining \li{read_data}:
% \begin{enumerate}
% \item The language's syntax definition determines that \li{read_data(a)} is an expression of function application form.
% \item Similarly, \li{read_data} and \li{a} are definitively expressions of variable form.
% \item The variable \li{a} can only refer to the binding of \li{a} on Line 1.
% \item The variable \li{w} can be renamed without knowing anything about the values that \li{read_data} and \li{a} stand for.
% \item The type of \li{x} can be determined to be \li{B} by determining that the type of \li{read_data} is \li{A -> B} for some \li{A} and \li{B}, and checking that \li{a} has type \li{A}. Nothing else needs to be known about the values that \li{read_data} and \li{a} stand for. In Reynolds' words \cite{B304}:
% \begin{quote}
% \emph{Type structure is a syntactic discipline for enforcing levels of abstraction.}
% \end{quote}
% \end{enumerate}

>>>>

% In short, the problem is that library providers cannot reason modularly about syntactic determinism (because other extensions might define overlapping forms), and if the desugaring of the program text is held abstract, programmers can no longer reason about types, binding and segmentation (i.e. answer questions like those above.) This is burdensome at all scales, but particularly when programming in the large, where it is common to encounter a wide variety of libraries \cite{DBLP:conf/sac/LammelPS11}. 
Forcing the programmer to reason transparently to answer basic questions like these defeats the ultimate purpose of syntactic sugar: decreasing cognitive cost \cite{Green89}. analogous problems do not arise when programming without syntax extensions in languages like ML --- programmers can reason lexically about where variables and other symbols are bound, and types mediate abstraction over function and module implementations \cite{B304}. Ideally, the programmer would be able to abstract in some analogous manner over the implementation of an unfamiliar notation. % From this perspective, it is unsurprising that the use of Camlp4 has been deprecated in the OCaml system.

@@

\subsubsection{Expansion Typing}
Having found the definition of the \li{$regex} TLM, the client can immediately determine the type of the expansion being generated at each application site because it is specified explicitly by the clause \li{at Regex.t} on Line~2 of Fig.~\ref{fig:regex-tlm-def}. The expansion type of a TLM is analogous to the return type of a function. The identity of \li{Regex.t} is determined relative to the TLM definition site, not at each application site, so the module \li{Regex} need not be in scope at the application site, or it can have been shadowed by a different module. % In Reason, TLM definitions can refer to exactly those modules that a Reason module at the top level of the defining package would have access to, i.e. other top-level modules in the package as well as modules exported by other packages that have been imported by the build system.

@@

Segment types are somewhat analogous to the argument types of a function. The difference is that the argument signature of a function is the same every time the function is applied, i.e. it is associated with the function itself, whereas the inferred segmentation can differ for each choice of literal body. This, of course, gives TLMs substantially more notational flexibility, even relative to infix or mixfix function notation (where the number of subexpressions is fixed \cite{wieland2009parsing}), as intended.% (even infix or mixfix functions \cite{wieland2009parsing}).

@@

% \subsection{Splicing}\label{sec:splicing-and-hygiene}
% When the parse function determines that some segment of the literal body is a spliced expression, according to whatever syntactic criteria it deems suitable, it can indirectly refer to it in the encoding it produces using the \li{SplicedE} constructor of \li{proto_expr}, which takes a value of type \li{segment} that indicates the zero-indexed location of the spliced expression relative to the start of the provided literal body. The \li{SplicedE} constructor also requires a value of type \li{proto_typ}, which indicates the type that the spliced expression is expected to have. Types can be spliced out by using the \li{SplicedT} constructor of \li{proto_typ} analogously.

@@

\begin{figure}[t]
\begin{subfigure}[t]{0.55\textwidth}
\begin{lstlisting}[morekeywords={expression}]
  module Lambda = {/* a typical encoding */};
  module LambdaNotation {
    notation $term at Lambda.term { 
      lexer SimpleLangLexer;
      expression parser LambdaParser.term_e;
      pattern parser LambdaParser.term_p;
      dependencies 
        { module Lambda = Lambda } };
    notation $v at Lambda.v {/* analogous */}
  };
\end{lstlisting}
\vspace{-6px}
\caption{Expression and pattern TLMs for \lismall{Lambda} terms and values}
\label{fig:lam}
\vspace{2px}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.42\textwidth}
\begin{lstlisting}
  open LambdaNotation;
  exception Unbound(Lambda.var);
  let rec eval = $term.(fun 
  | `(x)` => raise Unbound(x)
  | `(SURLlamEURL xSURL.EURLe)` => $v `(SURLlamEURL xSURL.EURLe)`
  | `(e1SURL(EURLe2SURL)EURL)` => {
    let $v `(SURLlamEURL xSURL.EURLe)` = eval(e1);
    let v2 = eval(e2);
    eval(`(SURL[EURLv2SURL/EURLxSURL]EURLe)`) 
  });
\end{lstlisting}
\vspace{-6px}
\caption{A reasonably elegant \lismall{Lambda} evaluator}
\label{fig:lam-eval}
\end{subfigure}
\vspace{2px}
\caption{Expression and pattern literal notation for lambda terms and values}
\vspace{-8px}
\label{fig:lam-example}
\end{figure}

@@

From a client programmer's perspective, reasoning principles analogous to those described in Sec. \ref{sec:client-perspective} for expression TLMs are available. \textbf{Responsibility} is assigned by the same protocol, and the type annotation on the responsible TLM governs the type of the generated pattern, satisfying the \textbf{Expansion Typing} condition. Patterns can contain module paths and type annotations, so pattern TLMs are governed by the same \textbf{Context Independence} condition as expression TLMs. The same protocol around \textbf{Segmentation} is also enforced.

@@

% \begin{figure}
% \begin{subfigure}[t]{0.51\textwidth}
% \begin{lstlisting}
%   module Lambda = {
%     type var = int;
%     type tm = Var(var) | Lam(var, tm) | Ap(tm, tm);
%     type v  = VLam(var, tm);
%     let subst : v -> var -> tm -> tm = (* ... *)
%   };
% \end{lstlisting}
% \caption{Lambda calculus terms}
% \end{subfigure}
% \hfill
% \begin{subfigure}[t]{0.49\textwidth}
% \begin{lstlisting}
%   module Lam = { /* a typical encoding */ };
%   module LamNotation {
%     notation $term at Lam.term { 
%       lexer SimpleLangLexer;
%       expression parser LamParser.tm_expr;
%       pattern parser Lam.tm_pat;
%       dependencies { module Lam }
%     }
%     notation $v at Lam.v { /* analogous */ } 
%   };
% \end{lstlisting}
% \vspace{-4px}
% \caption{Lambda TLM definitions}
% \vspace{2px}
% \end{subfigure}
% \caption{Lambda calculus example}
% \end{figure}

@@

% To maintain the abstract binding discipline, variable patterns can appear only within spliced patterns. Enforcing this restriction is straightforward: we simply have not defined a variant of the \li{proto_pat} type that encodes variable patterns (wildcards are allowed.)  This restriction ensures that only variables visible to the client in a spliced pattern are bound in the corresponding branch expression. This is analogous to the capture avoidance principle for expression TLMs.

@@

Instead, we define \emph{unexpanded typing contexts}, $\uGamma$, as pairs of the form $\uGG{\uG}{\Gamma}$, where $\uG$ maps each expression identifier $\ux \in \domof{\uG}$ to a variable, $x$, written $\vExpands{\ux}{x}$. The typing context, $\Gamma$, only tracks the type of these variables. We define the identifier update operation $\ctxUpdate{\uG}{\ux}{x}$ as producing the expression identifier expansion context that maps $\ux$ to $x$, written $\vExpands{\ux}{x}$, and defers to $\uG$ for all other expression identifiers, with no requirement that $\ux$ be apart from $\domof{\uG}$. %Note the distinction between update and extension (which requires that the new identifier is not already in the domain.) %We write $\uGammaOK{\uGamma}$ when $\uGamma=\uGG{\uG}{\Gamma}$ and each expression variable in $\uG$ is assigned a type by $\Gamma$.
%\begin{definition} $\uGammaOK{\uGG{\uG}{\Gamma}}$ iff for each $\vExpands{\ux}{x} \in \uG$, we have $\Ghyp{x}{\tau} \in \Gamma$ for some $\tau$.\end{definition}
%\noindent 
We define $\uGamma, \uGhyp{\ux}{x}{\tau}$ when $\uGamma = \uGG{\uG}{\Gamma}$ as an abbreviation of $\uGG{\ctxUpdate{\uG}{\ux}{x}}{\Gamma, \Ghyp{x}{\tau}}$. Unexpanded type formation contexts, $\uDelta$, are analogous.

@@

% \emph{Unexpanded type formation contexts}, $\uDelta$, consist of a \emph{type identifier expansion context}, $\uD$, paired with a standard type formation context, $\Delta$, and operate analogously (see supplement.)
% of the form $\uDD{\uD}{\Delta}$, i.e. they . We similarly define $\uDelta, \uDhyp{\ut}{t}$ when $\uDelta=\uDD{\uD}{\Delta}$ as an abbreviation of $\uDD{\ctxUpdate{\uD}{\ut}{t}}{\Delta, \Dhyp{t}}$.%type identifier expansion context is always extended/updated together with

@@

Rule \textsc{ee-ap-sptlm} for pattern TLM definitions shown in Fig. \ref{fig:ee-def-setlm} is analogous but simpler, because in $\miniVersePat$ patterns are entirely structural (there are no module paths that they might depend on). %As such, we do not need the machinery for dependency passing.
% {\vspace{-3px}\small\begin{mathpar}
% \inferrule[ee-def-sptsm]{
%   \expandsTU{\uDelta}{\utau}{\tau}\\
%   \hastypeU{\emptyset}{\emptyset}{\eparse}{\aparr{\tBody}{\tParseResultPat}}\\\\
%   \evalU{\eparse}{\eparse'}\\
%   \expandsUP{\uDelta}{\uGamma}{\uPsi}{\uPhi, \uPhyp{\tsmv}{a}{\tau}{\eparse'}}{\ue}{e}{\tau'}
% }{
%   \expandsUPX{\usyntaxup{\tsmv}{\utau}{\eparse}{\ue}}{e}{\tau'}
% }
% \end{mathpar}}
% \vspace{-3px}

@@

The typed pattern expansion rule governing pattern TLM application, Rule~\textsc{pe-ap-sptlm} in Fig.~\ref{fig:ee-ap-setlm}, is analogous but again simpler because there is no need to pass dependencies into patterns in $\miniVersePat$.\footnote{If there were a need to pass dependencies into patterns, we would need either pattern functions or to use an encoding trick, e.g. by requiring the result be a function containing a ``dummy'' match with one pattern.} We describe proto-pattern validation in Sec.~\ref{sec:s-PEV} below.
% {\small\begin{mathpar}
% \inferrule[pe-ap-sptsm]{
%   \encodeBody{b}{\ebody}\\
%   \evalU{\ap{\eparse}{\ebody}}{\aein{\mathtt{SuccessP}}{\ecand}}\\
%   \decodeCEPat{\ecand}{\cpv}\\\\
%   \segOK{\segof{\cpv}}{b}\\
%   \cvalidP{\upctx}{\pscene{\uDelta}{\uPhi}{b}}{\cpv}{p}{\tau}
% }{
%   \patExpands{\upctx}{\uPhi', \uPhyp{\tsmv}{a}{\tau}{\eparse}}{\utsmap{\tsmv}{b}}{p}{\tau}
% }
% \end{mathpar}}

@@

The rule for references to spliced unexpanded types and patterns are analogous (see supplement). Note that we did not give examples of type splicing in the previous sections, but it is occasionally useful. For example, a TLM that implements a parser generator in the style of Menhir could need type splicing for reading in the non-terminal types, e.g. as on Line 9 of Fig.~\ref{fig:regex-parser}.
% \noindent
% We first splice out the requested segment. The second premise expands the type annotation under an empty context, because the type annotation must be meaningful at the application site (so, independent of $\Delta$ and $\Gamma$) and not itself make any assumptions about the application site context. Spliced types can appear in the annotation. The third premise performs typed expansion of the spliced unexpanded expression under the application site contexts, but not the expansion-local contexts. The final two premises ensure that these contexts are disjoint, again to force capture avoidance.

% The rule for references to spliced unexpanded patterns is entirely analogous (see supplement).
% {\small\begin{mathpar}
% \inferrule[ppv-spliced]{
%   \parseUPat{\bsubseq{b}{m}{n}}{\upv}\\
%   \cvalidT{\emptyset}{\tsceneUP{\uDelta}{b}}{\ctau}{\tau}\\
%   \patExpands{\upctx}{\uPhi}{\upv}{p}{\tau}
% }{
%   \cvalidP{\upctx}{\pscene{\uDelta}{\uPhi}{b}}{\acesplicedp{m}{n}{\ctau}}{p}{\tau}
% }
% \end{mathpar}}

@@

% \endgroup
The additional second clause simply states that the final expansion produced by proto-expression validation is well-typed under the combined application site and expansion-internal context (because spliced terms are distinguished only in the proto-expansion, but not in the final expansion). % The combined context can only be formed if these are disjoint.
The proof proceeds by mutual rule induction and appeal to the analogous typed pattern expansion theorem and simple lemmas about type expansion   and proto-type validation. The only issue is that it is not immediately clear that the mutual induction is well-founded, because the case in the proof of part 2 for Rule \textsc{pev-spliced} invokes part 1 of the induction hypothesis on a term that is not a sub-term of the conclusion, but rather parsed out of the literal body, $b$. To establish that the mutual induction is well-founded, then, we need to explicitly establish a decreasing metric. The intuition is that parsing a term of out a literal body cannot produce a bigger term than the term that contained that very literal body. 
% More specifically, {the sum of the lengths of the literal bodies that appear in the term strictly decreases each time you perform a nested TLM application} because some portion of the term has to be consumed by the TLM name and the delimiters. 
The details are given in the supplement.

@@

% % The following theorem summarizes the abstract reasoning principles available to programmers when applying a pattern TLM. Most of the labeled clauses are analogous to those described above, so we omit their descriptions.
% % % \begingroup
% % % \def\thetheorem{\ref{thm:spTLM-Typing-Segmentation}}
% % \begin{theorem}[spTLM Reasoning Principles]
% % % \label{thm:spTLM-Typing-Segmentation}
% % If $\patExpands{\upctx}{\uPhi}{\utsmap{\tsmv}{b}}{p}{\tau}$ where $\uDelta=\uDD{\uD}{\Delta}$ and $\uGamma=\uGG{\uG}{\Gamma}$ then all of the following hold:
% % \begin{enumerate}[noitemsep,nolistsep]
% %         \item (\textbf{Typing 1}) $\uPhi=\uPhi', \uPhyp{\tsmv}{a}{\tau}{\eparse}$ and $\patType{\pctx}{p}{\tau}$
% %         \item $\encodeBody{b}{\ebody}$
% %         \item $\evalU{\eparse(\ebody)}{\aein{\mathtt{SuccessP}}{\ecand}}$
% %         \item $\decodeCEPat{\ecand}{\cpv}$
% %         \item (\textbf{Segmentation}) $\segOK{\segof{\cpv}}{b}$
% %         \item $\segof{\cpv} = \sseq{\acesplicedt{n'_i}{m'_i}}{\nty} \cup \sseq{\acesplicedp{m_i}{n_i}{\ctau_i}}{\npat}$
% %         \item (\textbf{Typing 2}) $\sseq{
% %               \expandsTU{\uDelta}
% %               {
% %                 \parseUTypF{\bsubseq{b}{m'_i}{n'_i}}
% %               }{\tau'_i}
% %             }{\nty}$ and $\sseq{\istypeU{\Delta}{\tau'_i}}{\nty}$
% %         \item (\textbf{Typing 3}) $\sseq{
% %           \cvalidT{\emptyset}{
% %             \tsceneUP
% %               {\uDelta}{b}
% %           }{
% %             \ctau_i
% %           }{\tau_i}
% %         }{\npat}$ and $\sseq{\istypeU{\Delta}{\tau_i}}{\npat}$
% %         \item (\textbf{Typing 4}) $\sseq{
% %           \patExpands
% %             {\uGG{\uG_i}{\pctx_i}}
% %             {\uPhi}
% %             {\parseUPatF{\bsubseq{b}{m_i}{n_i}}}
% %             {p_i}
% %             {\tau_i}
% %         }{\npat}$  and $\sseq{\patType{\pctx_i}{p_i}{\tau_i}}{\npat}$
% %       \item (\textbf{Visibility}) $\uG = \biguplus_{0 \leq i < \npat} \uG_i$ and $\Gamma = \bigcup_{0 \leq i < \npat} \pctx_i$
% %         \begin{quote}
% %           \begin{grayparbox}
% %           The hypotheses generated by the TLM application are exactly those generated by the spliced patterns.
% %           \end{grayparbox}
% %         \end{quote}
% % \end{enumerate}
% % \end{theorem}
% % 
% % \begin{proof} 
% % The proof, in the supplement, relies on an auxiliary lemma about decomposing proto-patterns.

@@

% % % \item[{\text{(\ref{rule:cvalidE-U-tap})}}~\textbf{through}~{\text{(\ref{rule:cvalidE-U-case})}}] These cases follow analogously, i.e. we apply the IH, part 2 to all proto-expression validation judgements, Lemma \ref{lemma:candidate-expansion-type-validation} to all proto-type validation judgements, the identification convention to ensure that extended contexts remain disjoint, weakening and exchange as needed, and the corresponding typing rule in Rules (\ref{rule:hastypeU-tap}) through (\ref{rule:hastypeU-case}).
% % % \\

@@

% % % \item[\VExpof{\text{\ref{rule:hastypeU-tlam}}}~\text{through}~\VExpof{\text{\ref{rule:hastypeU-case}}}] These cases follow analogously, i.e. we apply the IH to all proto-expression validation premises, Lemma \ref{lemma:candidate-expansion-type-validation} to all proto-types validation premises, weakening and exchange as needed, and then apply the corresponding typing rule.
% % % \\

@@

% In contrast, when a programmer encounters, for example, a function call like the call to \li{read_data} on Line 3, the analogous questions can be answered by following clear protocols that become ``cognitive reflexes'' after sufficient experience with the language, even if the programmer has no experience with the library defining \li{read_data}:
% \begin{enumerate}
% \item The language's syntax definition determines that \li{read_data(a)} is an expression of function application form.
% \item Similarly, \li{read_data} and \li{a} are definitively expressions of variable form.
% \item The variable \li{a} can only refer to the binding of \li{a} on Line 1.
% \item The variable \li{w} can be renamed without knowing anything about the values that \li{read_data} and \li{a} stand for.
% \item The type of \li{x} can be determined to be \li{B} by determining that the type of \li{read_data} is \li{A -> B} for some \li{A} and \li{B}, and checking that \li{a} has type \li{A}. Nothing else needs to be known about the values that \li{read_data} and \li{a} stand for. In Reynolds' words \cite{B304}:
% \begin{quote}
% \emph{Type structure is a syntactic discipline for enforcing levels of abstraction.}
% \end{quote}
% \end{enumerate}
